---
title: "rPraat package for R"
output: md_document
---

# rPraat package for R

rPraat package for R constitutes an interface between the most popular software for phonetic analyses, Praat, and R.

Version 1.1.3-2.

For demonstration, see [html/rDemo.html](http://htmlpreview.github.com/?https://github.com/bbTomas/rPraat/blob/master/html/rDemo.html)

Package documentation [https://github.com/bbTomas/rPraat/blob/master/rPraat.pdf](https://github.com/bbTomas/rPraat/blob/master/rPraat.pdf)

CRAN link: [https://CRAN.R-project.org/package=rPraat](https://CRAN.R-project.org/package=rPraat)

See also: [mPraat toolbox](https://github.com/bbTomas/mPraat) for Matlab.

rPraat and mPraat homepage: [http://fu.ff.cuni.cz/praat/](http://fu.ff.cuni.cz/praat/)

## Installation

One-time installation (internet connection needed). 

#### CRAN: official version

```{r, eval=FALSE}
install.packages("rPraat")
```

#### GitHub: experimental, most recent version

```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("bbTomas/rPraat")
```


### Initialize and Test

At each R session, you have to initialize the package (no internet needed).

```{r, eval=FALSE}
library(rPraat)
library(dplyr)   # optional, if you want to use the pipeline operator %>%

tg.sample() %>% tg.plot()  # demo test
```


## How to cite this package?
We have invested a lot of time and effort in creating mPraat toolbox / rPraat package, please cite it when using it for data analysis.

> Bořil, T., & Skarnitzl, R. (2016). Tools rPraat and mPraat. In P. Sojka, A. Horák, I. Kopeček, & K. Pala (Eds.), Text, Speech, and Dialogue (pp. 367–374). Springer International Publishing.

[Download Tools rPraat and mPraat manuscript .pdf](http://fu.ff.cuni.cz/praat/boril_skarnitzl_2016_Tools_rPraat_and_mPraat_[manuscript].pdf)

The final publication is available at Springer via [DOI 10.1007/978-3-319-45510-5_42](http://dx.doi.org/10.1007/978-3-319-45510-5_42)


[Download BibTeX reference mpraatrpraat2016.bib](http://fu.ff.cuni.cz/praat/mpraatrpraat2016.bib)

## What is new

v1.1.3-2 (2018/10/29): formant.plot(), formant.toArray(); tg.plot() can plot TextGrid together with Formant object

v1.1.3-1 (2018/10/26): tg.plot() can plot TextGrid together with PitchTier and IntensityTier in one plot

v1.1.3 (2018/10/20): formant.read(), formant.sample(), as.formant(); col.read() can read formant objects

v1.1.2-1 (2018/10/20): tg.plot(): right-aligned labels; pitch.sample(), as.tg(), as.pt(), as.it(), as.pitch(); .read(), .sample() and tg.createNewTextGrid(): class(object)["type"] set to "TextGrid", "PitchTier" etc. and class(object)["name"] set according to fileName

v1.1.2 (2018/08/04): tg.cut(), tg.cut0()

v1.1.1 (2018/08/02): IntensityTiers: it.read(), it.write(), it.plot(), it.sample(), it.interpolate(), it.cut(), it.cut0(), it.legendre(), it.legendreDemo(), it.legendreSynth(), col.read() supports IntensityTiers

v1.1.0 (2018/08/01): col.read() to read Collection in Text and Short-text format (Collection may store many objects in one file – TextGrids, PitchTiers, Pitch objects). Big thanks to Pol van Rijn for initiating this and making the major portion of the new code in mPraat toolbox. Supports for file encodings in all read functions (thanks to Weirong Chen for his DetectEncoding idea in mPraat toolbox).

v1.0.8-3 (2017/11/27): tg.write(): fix for possible UTF-8 encoding and Windows OS locale mismatch

v1.0.8-2 (2017/11/04): tg.write() and pt.write() format support: "text" (full text format), "short" (short text format), and in addition for pt: "spreadsheet" and "headerless"; tg.read(), pt.read() and pitch.read(): optional file encoding parameter (default: "UTF-8")

v1.0.8-1 (2017/09/01): bugfix, short-text TextGrid with multiple lines in a label no longer produces "missing sppasFormat variable" error

v1.0.8 (2017/07/16): pitch.read() to read Pitch-object files with time frames of pitch candidates

v1.0.7 (2017/04/19): tg.findLabels(), tg.duplicateTierMergeSegments()

v1.0.6 (2017/04/11): pt.cut(), pt.cut0()

v1.0.5 (2017/04/10): pt.legendre(), pt.legendreSynth(), pt.legendreDemo()

v1.0.4 (2017/03/06): pt.interpolate(), pt.Hz2ST()

## License

License: MIT

Copyright (c) 2016 Tomáš Bořil, borilt@gmail.com


Permission is hereby granted, free of charge, to any person obtaining a copy of this software and associated documentation files (the "Software"), to deal in the Software without restriction, including without limitation the rights to use, copy, modify, merge, publish, distribute, sublicense, and/or sell copies of the Software, and to permit persons to whom the Software is furnished to do so, subject to the following conditions:

The above copyright notice and this permission notice shall be included in all copies or substantial portions of the Software.

THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE SOFTWARE.
