# rPraat: Praat in R

## How to cite this package / toolbox?
We have invested a lot of time and effort in creating mPraat toolbox / rPraat package, please cite it when using it for data analysis.

> Bořil, T., & Skarnitzl, R. (2016). Tools rPraat and mPraat. In P. Sojka, A. Horák, I. Kopeček, & K. Pala (Eds.), Text, Speech, and Dialogue (pp. 367–374). Springer International Publishing.

[Download Tools rPraat and mPraat manuscript .pdf](boril_skarnitzl_2016_Tools_rPraat_and_mPraat_[manuscript].pdf)

The final publication is available at Springer via [DOI 10.1007/978-3-319-45510-5_42](http://dx.doi.org/10.1007/978-3-319-45510-5_42)


[Download BibTeX reference mpraatrpraat2016.bib](mpraatrpraat2016.bib)


## Installation and help

rPraat and mPraat homepage: [http://fu.ff.cuni.cz/praat/](http://fu.ff.cuni.cz/praat/)

Package rPraat at github: [https://github.com/bbTomas/rPraat/](https://github.com/bbTomas/rPraat/)

```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("bbTomas/rPraat")
```

At each R session, you have to initialize the package.
```{r, eval=FALSE}
library(rPraat)
library(dplyr)   # optional, if you want to use the pipeline operator %>%
```


For help and examples, use command
```{r, eval=FALSE}
?nameOfFunction
```
or read .pdf documentation at [https://github.com/bbTomas/rPraat/blob/master/rPraat.pdf](https://github.com/bbTomas/rPraat/blob/master/rPraat.pdf)


## Demo files
To reproduce the following examples, you will need some sample files. Download and extract [rpraat_demo.zip](http://fu.ff.cuni.cz/praat/rpraat_demo.zip) to your Working Directory, it creates `demo` subdirectory.

### Read and plot Sound, TextGrid, and PitchTier
```{r, fig.height=2.5}
library(tuneR)
library(tbTools)
library(dygraphs)
library(dplyr)

sndWav <- readWave("demo/H.wav"); fs <- sndWav@samp.rate; snd <- sndWav@left / (2^(sndWav@bit-1))
t <- seqM(0, (length(snd)-1)/fs, by = 1/fs)

dygraph(data.frame(t, snd), xlab = "Time (sec)", group="group1") %>% dyRangeSelector() %>% dyAxis("y", label = "Amplitude")
```

```{r, fig.height=3}
library(rPraat)
library(dplyr)
tg <- tg.read("demo/H.TextGrid") %>%
    tg.removeTier("phrase") %>% tg.removeTier("phoneme") %>% tg.removeTier("syllable")

tg.plot(tg, group="group1")
```

```{r, fig.height=2.5}
pt <- pt.read("demo/H.PitchTier")
pt.plot(pt, group="group1") %>% dyAxis("y", label = "Frequency (Hz)")
```

#### Interactive plots using dygraphs (including tg.plot, pt.plot)

Mouse drag: Zoom to selection

Mouse Double-click: Default zoom

Shift + mouse drag: plot scroll


## TextGrid

```{r}
tg <- tg.read("demo/H.TextGrid")
tg.getNumberOfTiers(tg)
length(tg)   # number of tiers, "low-level access"
tg.getTotalDuration(tg)
tg.getTierName(tg, 1)
```

### Tier accessed both by index and name

```{r}
tg.isPointTier(tg, 1)
tg.isPointTier(tg, "phoneme")

tg.isIntervalTier(tg, 1)
tg.isIntervalTier(tg, "phoneme")

tg[[1]]$type
tg$phoneme$type

tg.checkTierInd(tg, "word")  # get index of a tier with specified name
tg.checkTierInd(tg, 4)  # only checks whether TextGrid contains a tier with this index
```

### Point tier

```{r}
tg.getNumberOfPoints(tg, 1)
tg.getNumberOfPoints(tg, "phoneme")
tg.getPointTime(tg, "phoneme", 4)
tg.getLabel(tg, "phoneme", 4)
```

#### Point tier "low-level access"

```{r}
length(tg[[1]]$t)   # number of points
tg[[1]]$t[4]
tg[[1]]$label[4]

length(tg$phoneme$t)
tg$phoneme$t[4]
tg$phoneme$label[4]

tg$phoneme$t[5:8]
```

### Interval tier

```{r}
tg.isPointTier(tg, "word")
tg.isIntervalTier(tg, "word")
tg[[4]]$type
tg$word$type

tg.getTotalDuration(tg, "word")
tg.getStartTime(tg, "word")
tg.getEndTime(tg, "word")

tg.getNumberOfIntervals(tg, "word")
tg.getIntervalStartTime(tg, "word", index = 4)
tg.getIntervalEndTime(tg, "word", index = 4)
tg.getIntervalDuration(tg, "word", index = 4)
tg.getLabel(tg, "word", index = 4)
```

#### Interval tier "low-level access"

```{r}
length(tg[[4]]$t1)   # number of intervals
tg[[4]]$t1[4]
tg[[4]]$t2[4]
tg[[4]]$label[4]

length(tg$word$t1)
tg$word$t1[4]
tg$word$t2[4]
tg$word$label[4]

tg$word$label[5:8]
```

### Vectorized operations

```{r, fig.height=3.5}
labelsOfInterest <- c("i", "i:", "e", "e:", "a", "a:", "o", "o:", "u", "u:")
condition <- tg$phone$label %in% labelsOfInterest

sum(condition) # How many labels satisfy the condition?

dur <- tg$phone$t2[condition] - tg$phone$t1[condition]
mean(dur)
hist(dur)
```

### Find labels in TextGrid

```{r}
tg <- tg.read("demo/H.TextGrid")

# find label "n" in phoneme tier (Point tier)
i <- tg.findLabels(tg, "phoneme", "n")
i # 4 results and their indices
length(i) # 4 results
i[[1]]  # index of the first result
tg$phoneme$label[unlist(i)]

# find fragments with successive labels "?" "a" in phone tear (Interval tier)
i <- tg.findLabels(tg, "phone", c("?", "a"))
i  
length(i) # 2 results found
i[[1]]  # indices of the first result
i[[2]]  # indices of the second result
tg$phone$label[i[[1]]]
tg$phone$label[i[[2]]]
tg$phone$label[unlist(i)]

# find fragments with successive labels "?" "a" in phone tier
# and return initial and final time of these fragments
t <- tg.findLabels(tg, "phone", c("?", "a"), returnTime = TRUE)
t # 2 results and their initial (t1) and final (t2) time
length(t) # 2 results found
t$t2[1] - t$t1[1]   # duration of the first fragment
t$t2[2] - t$t1[2]   # duration of the second fragment

# find fragments with successive labels "ti" "řeknu" "co" in word tier (Interval tier)
i <- tg.findLabels(tg, "word", c("ti", "řeknu", "co"))
i  
length(i) # 1 result found
i[[1]]           # indices of segments in the first result
length(i[[1]])   # length of the first (and only) result (number of segments)
i[[1]][3]        # index of the third segment in the first result
tg$word$label[i[[1]]]  # all labels in the fragment
# get initial and final time of the fragment
t <- tg.findLabels(tg, "word", c("ti", "řeknu", "co"), returnTime = TRUE)
t
```

### Create syllable tier from phone tier

```{r}
tg <- tg.read("demo/H.TextGrid")

tg <- tg %>% tg.removeTier("phoneme") %>%  tg.removeTier("syllable") %>% tg.removeTier("word") %>%
    tg.removeTier("phrase")
```

Get actual labels.

```{r}
collapsed <- paste0(tg$phone$label, collapse = "")
collapsed
```

Edit the collapsed string with labels - insert separators to mark boundaries of syllables.

* There can be segments with empty labels in the original tier (pauses), do not specify them in the pattern.
* Beware of labels that appear empty but they are not (space, new line character etc.) - these segments are handled as classical non-empty labels. See example - one label is " ", therefore it must be specified in the pattern.

```{r}
pattern <- "ja:-ci-P\\ek-nu-t_so-?u-J\\e-la:S- -nej-dP\\i:f-naj-deZ-h\\ut_S-ku-?a-?a-ta-ma-na:"
tg2 <- tg.duplicateTierMergeSegments(tg, "phone", 1, "syll", pattern, sep = "-")
tg.plot(tg2)
```



### Overview of some TextGrid operations

For all functions, see help `?functionName` for its description and example of use.

```{r, fig.height=3.5}
tg <- tg.read("demo/H.TextGrid")
tg.plot(tg)

tg <- tg.removeTier(tg, "syllable")
tg <- tg.removeTier(tg, "phrase")
tg <- tg.removeTier(tg, "phone")

ind <- tg.getPointIndexNearestTime(tg, "phoneme", time = 1.5)
tg <- tg.setLabel(tg, "phoneme", ind, newLabel = "!Q!")

tg <- tg.insertPoint(tg, "phoneme", time = 1.6, label = "NEW")
tg$phoneme$t <- tg$phoneme$t[-(30:40)]   # remove points 30:40
tg$phoneme$label <- tg$phoneme$label[-(30:40)]

tg <- tg.duplicateTier(tg, "word", newInd = 2, newTierName = "WORD2")
tg <- tg.removeIntervalBothBoundaries(tg, "WORD2", index = 6)
tg <- tg.setLabel(tg, "WORD2", index = 5, newLabel = "")
tg <- tg.insertInterval(tg, "WORD2", tStart = 0.9, tEnd = 1.7, label = "NEW LAB")
ind <- tg.getIntervalIndexAtTime(tg, "WORD2", time = 2.3)
tg <- tg.removeIntervalBothBoundaries(tg, "WORD2", ind)
tg.plot(tg)

tgNew <- tg.createNewTextGrid(tMin = 0, tMax = 5)
tgNew <- tg.insertNewIntervalTier(tgNew, newInd = 1, "word")   # the first tier
tgNew <- tg.insertInterval(tgNew, tierInd = 1, tStart = 2, tEnd = 3.5, "hello")
tgNew <- tg.insertInterval(tgNew, tierInd = 1, 4, 4.8, "world")
tgNew <- tg.insertNewIntervalTier(tgNew, newInd = Inf, "word_last")  # the last tier (at this moment)
tgNew <- tg.insertInterval(tgNew, tierInd = "word_last", 2, 3, "ABC")
tgNew <- tg.insertNewPointTier(tgNew, newInd = 2, newTierName = "click")
tgNew <- tg.insertPoint(tgNew, tierInd = 2, time = 2, label = "click")
tgNew <- tg.insertPoint(tgNew, tierInd = 2, time = 4, label = "click")
tgNew <- tg.insertNewPointTier(tgNew, newInd = Inf, newTierName = "pointTierLast")
tgNew <- tg.insertPoint(tgNew, tierInd = "pointTierLast", time = 3, label = "point")
tg.plot(tgNew)
tg.write(tgNew, "demo/ex_output.TextGrid")
```



### Repair continuity problem of TextGrid

Repairs problem of continuity of T2 and T1 in interval tiers. This problem is very rare and it should not appear. However, e.g., automatic segmentation tool Prague Labeller produces random numeric round-up errors featuring, e.g., T2 of preceding interval is slightly higher than the T1 of the current interval. Because of that, the boundary cannot be manually moved in Praat edit window.

```{r}
tgProblem <- tg.read("demo/H_problem.TextGrid")
tgNew <- tg.repairContinuity(tgProblem)
tg.write(tgNew, "demo/H_problem_OK.TextGrid")

tgNew2 <- tg.repairContinuity(tgNew) # no problem in repaired TextGrid
```

## PitchTier

**Example.** Transform Hz to semitones (ST), cut the original pitchtier along the TextGrid, make interpolated contour.

```{r, fig.height=3.5}
pt <- pt.read("demo/H.PitchTier")
pt <- pt.Hz2ST(pt, ref = 100)  # conversion of Hz to Semitones, reference 0 ST = 100 Hz.

pt.plot(pt) %>% dygraphs::dyAxis("y", label = "Frequency (ST re 100 Hz)")

tg <- tg.read("demo/H.TextGrid")
t <- tg.findLabels(tg, "word", c("já", "ti", "řeknu"), returnTime = TRUE)
print(t)

pt2 <- pt.cut0(pt, tStart = t$t1[1], tEnd = t$t2[1]) # "já ti řeknu"
pt.plot(pt2) %>% dygraphs::dyAxis("y", label = "Frequency (ST re 100 Hz)")

pt2interp <- pt.interpolate(pt2, seq(pt2$t[1], pt2$t[length(pt2$t)], by = 0.001))
pt.plot(pt2interp) %>% dygraphs::dyAxis("y", label = "Frequency (ST re 100 Hz)")

pt.write(pt2interp, "demo/H_cut_interp.PitchTier")
```

### Legendre polynomials modelling


#### Orthogonal basis
```{r}
pt.legendreDemo()
```

#### Cut the PitchTier
```{r}
pt <- pt.sample()
pt <- pt.Hz2ST(pt)
pt <- pt.cut(pt, tStart = 3)  # cut PitchTier from t = 3 sec and preserve time
```

#### Model it using Legendre polynomials
```{r}
c <- pt.legendre(pt)
print(c)
```
#### Reconstruct the contour from these 4 coefficients
```{r, fig.height=5}
leg <- pt.legendreSynth(c)
ptLeg <- pt
ptLeg$t <- seq(ptLeg$tmin, ptLeg$tmax, length.out = length(leg))
ptLeg$f <- leg

plot(pt$t, pt$f, xlab = "Time (sec)", ylab = "F0 (ST re 100 Hz)")
lines(ptLeg$t, ptLeg$f, col = "blue")
```

## Process all files in folder

```{R, eval = FALSE}
inputFolder <- "experiment1/data"
listFiles <- list.files(path = inputFolder, pattern = "\\.TextGrid$" , ignore.case = TRUE)

for (I in seq_along(listFiles)) {
    file <- listFiles[I]
    fileName <- substr(file, 1, nchar(file)-9)
    fileTextGrid <- paste0(inputFolder, "/", fileName, ".TextGrid")
    filePitchTier <- paste0(inputFolder, "/", fileName, ".PitchTier")
    fileSound <- paste0(inputFolder, "/", fileName, ".wav")
    
    tg <- tg.read(fileTextGrid)
    pt <- pt.read(filePitchTier)
    # ...
}

```










